Carrier phase and amplitude estimation for phase shift keying using pilots and data

### 2014-05-26 paper.tex
added further to comments

In this paper we present an alternative approach to the study of estimators for carrier phase and amplitude.  Rather than compute bounds, we directly study the least squares estimator asymptotically as the number of transmitted symbols $L\to \infty$.  We show, under comparatively mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent (asymptotically unbiased) and asymptotically normally distributed.  However, the least squares amplitude estimator is asymptotically biased and converges to a positive real number that is a function of the true carrier amplitude, the noise distribution, and the size of the constellation $M$.  To our knowledge, this is the first time an analytic description of the bias or variance of a non-data-aided amplitude estimator has been obtained.  Our expressions for the asymptotic variance and bias are simple to compute.  %Our results are indirectly supported by the simulation study of Pauluzzi~\cite{Pauluzzi2000}.  We present our our set of simulations supporting our analytic results in Section~\ref{}.

@@

Turbo synchronisation can require a large number of EM iterations.  For applications with low power receivers, such as sensor networks, the complexity of turbo synchronisation might be too great and computationally simple channel estimators are required.  For example, in practice it is common to rely on channel estimates obtained from pilot symbols alone.  In this case the number of pilots symbols must be sufficiently large for channel estimates to be sufficiently accurate.  If more accuracy is required, then more pilots symbols must be used.  Our estimator provides a simple mechanism for increasing accuracy without increasing the number pilot symbols.  Instead, the pilots symbols and a subset of the data symbols can be jointly used for channel estimation.  %For computational complexity reasons it may not be desirable to use all of the existing data symbols for channel estimation.  This is easily possible with our estimator by choosing the indices in $D$ to correspond with the subset of the data symbols that are to be used.

@@

Define the sum of squares function
\begin{align}
LS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \nonumber \\
&= \sum_{i \in P \cup D} \big( \abs{y_i}^2 - a s_i y_i^* - a^* s_i^* y_i + aa^* \big), \label{eq:LS}
\end{align}
where $*$ denotes complex conjugate.  The minimiser of $LS$ with respect to $a$ as a function of $\{d_i, i \in D\}$ is
\begin{equation}\label{eq:hata}
\hat{a}(\{d_i, i \in D\}) = \frac{1}{L} \sum_{i \in P \cup D} y_i s_i^* = \frac{1}{L} Y,
\end{equation}
where $L = \abs{P \cup D}$ and to simplify notation we have put 
\[
Y = \sum_{i \in P \cup D} y_i s_i^* = \sum_{i \in P } y_i p_i^* + \sum_{i \in D } y_i d_i^*.
\]  
Note that $Y$ is a function of the unknown data symbols $\{ d_i, i \in D\}$ and we could write $Y(\{ d_i, i \in D\})$, but have chosen to suppress the argument $(\{ d_i, i \in D\})$ for notational brevity.  Substituting $\frac{1}{L}Y$ for $a$ into~\eqref{eq:LS} we obtain $LS$ minimised with respect to $a$,
\begin{equation}\label{eq:LSdatasymbols}
LS(\{d_i, i \in D\}) = A - \frac{1}{L}\abs{Y}^2,
\end{equation}
where $A = \sum_{i \in P \cup D}\abs{y_i}^2$ does not depend on the data symbols.  The least squares estimators of the data symbols are the minimisers of~\eqref{eq:LSdatasymbols}.  Observe that given candidate values for the data symbols, we can compute the corresponding $LS(\{d_i, i \in D\})$ in $O(L)$ arithmetic operations.  It turns out that there are at most $M\abs{D}$ candidate values of the least squares estimator of the data symbols~\cite{Sweldens2001,Mackenthun1994}.  %When the number of data symbols is not small, this set is substantially smaller than the entire set of possible transmitted symbols, which is of size $M^{\abs{D}}$.

To see this, let $a = \rho e^{j\theta}$ where $\rho > 0$ is real.  Now $LS$ from~\eqref{eq:LS} is to
\begin{align}
LS(\rho, \theta, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - \rho e^{j\theta} s_i }^2  \nonumber \\
&= \sum_{i \in P} \abs{ y_i - \rho e^{j\theta} p_i }^2 + \sum_{i \in D} \abs{ y_i - \rho e^{j\theta} d_i }^2. \label{eq:LSallparams}
\end{align}
We have slightly abused notation here by reusing $LS$. This should not cause confusion as $LS(a, \{d_i, i \in D\})$, $LS(\rho, \theta, \{d_i, i \in D\})$, and $LS(\{d_i, i \in D\})$ are easily told apart by their arguments.  For given $\theta$, the least squares estimator of the $i$th data symbol $d_i$ is given by minimising $\abs{ y_i - \rho e^{j\theta} d_i }^2$, that is,
\begin{equation}\label{eq:hatdfinxtheta}
\hat{d}_i(\theta) = e^{j\hat{u}_i(\theta)} \qquad \text{where} \qquad \hat{u}_i(\theta) = \round{\angle( e^{-j\theta}y_i)},
\end{equation}
where $\angle(\cdot)$ denotes the complex argument (or phase), and $\round{\cdot}$ rounds its argument to the nearest multiple of $\frac{2\pi}{M}$.  A word of caution, the notation $\round{\cdot}$ is often used to denote rounding to the nearest \emph{integer}.  This is not the case here.  If the function $\operatorname{round}(\cdot)$ takes its argument to the nearest integer then,
\[
\round{x} = \tfrac{2\pi}{M}\operatorname{round}\left(\tfrac{M}{2\pi}x\right).
\] 
It is common to call $\hat{d}_i(\theta)$ a `hard decision'  made for the $i$th data symbol under the hypothesis that the channel phase is $\theta$.  Note that $\hat{d}_i(\theta)$ does not depend on $\rho$.  As defined, $\hat{u}_i(\theta)$ is not strictly inside the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$, but this is not of consequence, as we intend its value to be considered equivalent modulo $2\pi$.  With this in mind,
\[
\hat{u}_i(\theta) = \round{\angle{y_i} - \theta }
\]
which is equivalent to the definition from~\eqref{eq:hatdfinxtheta} modulo $2\pi$.

@@

Let,
\begin{equation}\label{eq:LSfk}
LS(f_k) = A - \frac{1}{L}\abs{Y_k}^2,
\end{equation}
where, 
\[
Y_k = Y( f_k ) = \sum_{i \in P} y_i p_i^*  + \sum_{i \in D} y_i f_{ki}^* = B + \sum_{i \in D}g_{ki},
\]
where $B = \sum_{i \in P} y_i p_i^*$ is independent of the data symbols, and $f_{ki}$ denotes the $i$th symbol in $f_k$, and for convenience, we put $g_{ki}  = y_i f_{ki}^*$.  Letting $g_{k}$ be the sequence $\{g_{ik}, i \in D\}$ we have, from~\eqref{eq:fkrec}, that $g_k$ satisfies the recursive equation
\[
g_{k+1} = g_{k} e_{\sigma(k)}^*,
\]
where $g_{k} e_{\sigma(k)}^*$ indicates the sequence $g_k$ with the $\sigma(k)$th element replaced by $g_{k \sigma(k)}e^{j2\pi/M}$.  Now,
\[
Y_0 = B + \sum_{i \in D} g_{0i}
\] 
can be computed in $O(L)$ operations, and
\begin{align*}
Y_1 &= B + \sum_{i \in D} g_{1i} \\
&= B +  (e^{j2\pi/M} - 1)g_{0\sigma(0)} + \sum_{i \in D} g_{0i} \\
&= Y_0 + \eta g_{0\sigma(0)},
\end{align*}
where $\eta = e^{j2\pi/M} - 1$.  In general,
\[
Y_{k+1} = Y_k + \eta g_{k\sigma(k)}.
\]
So, each $Y_k$ can be computed from its predecessor $Y_{k-1}$ in $O(1)$ arithmetic operations.  Given $Y_k$, the value of $LS(f_k)$ can be computed in $O(1)$ operations using~\eqref{eq:LSfk}.  Let $\hat{k} = \arg\min LS(f_k)$.  The least squares estimator of $a_0$ is then computed according to~\eqref{eq:hata},
\[
\hat{a} = \frac{1}{L} Y_{\hat{k}}.
\]
Pseudocode is given in Algorithm~\ref{alg:loglinear}.  Line~\ref{alg_sortindices} contains the function $\operatorname{sortindices}$ that, given $z = \{z_i, i \in D\}$, returns the permutation $\sigma$ as described in~\eqref{eq:sigmasortind}.  The $\operatorname{sortindices}$ function requires sorting $\abs{D}$ elements.  This requires $O(L \log L)$ operations.  The $\operatorname{sortindices}$ function is the primary bottleneck in this algorithm when $L$ is large.  The loops on lines~\ref{alg_loop_setup} and~\ref{alg_loop_search} and the operations on lines~\ref{alg_Y} to lines~\ref{alg_Q} all require $O(L)$ or less operations.  %In the next sections we will show how to the sortinces function can be avoided.  This leads to an algorithm that requires only $O(L)$ operations.

@@

\begin{theorem}\label{thm:consistency} (Almost sure convergence)
Let $\{w_i\}$ be a sequence of independent and identically distributed, circularly symmetric complex random variables with $\expect \abs{w_1}^2$ finite, and let $\{y_i, i \in P \cup D\}$ be given by~\eqref{eq:sigmod}.   Let $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ be the least squares estimator of $a_0 = \rho_0e^{j\theta_0}$. %where both $\hat{\rho}$ and $\rho_0$ are positive real numbers.  
Put $L = \abs{P \cup D}$ and let $\abs{P}$ and $\abs{D}$ increase in such a way that
\[
\frac{\abs{P}}{L} \rightarrow p \qquad \text{and} \qquad \frac{\abs{D}}{L} \rightarrow d \qquad \text{as $L \rightarrow \infty$.}
\] 
Let $R_i \geq 0$ and $\Phi_i \in [0,2\pi)$ be real random variables satisfying
\begin{equation}\label{eq:RiandPhii}
R_ie^{j\Phi_i} = 1 + \frac{w_i}{a_0 s_i} ,
\end{equation}
and define the continuous function
\[
G(x) = p h_1(x) + d h_2(x) \qquad \text{where}
\]
\[
h_1(x) = \expect R_1 \cos(x + \Phi_1), \;\;\; h_2(x) =  \expect R_1 \cos\sfracpart{ x + \Phi_1}.
\]
If $p > 0$ and if $G(x)$ is uniquely maximised at $x = 0$ over the interval $[-\pi,\pi)$ then
\begin{enumerate}
\item $\sfracpart{\theta_0 - \hat{\theta}}_\pi \rightarrow 0$ almost surely as $L \rightarrow \infty$, \label{thm:consistency:part1}
\item $\hat{\rho} \rightarrow \rho_0 G(0)$ almost surely as $L \rightarrow \infty$. \label{thm:consistency:part2}
\end{enumerate}
\end{theorem}

@@

The proof of Theorem~\ref{thm:consistency} is in Section~\ref{sec:proof-almost-sure} and the proof of Theorem~\ref{thm:normality} is in Section~\ref{sec:proof-asympt-norm}. Before giving the proofs we discuss some of the consequences and the assumptions made by these theorems.  
First observe that the theorems place conditions on $\sfracpart{\theta_0 - \hat{\theta}}_\pi$ rather than directly on $\theta_0 - \hat{\theta}$.  This makes sense because the phases $\theta_0$ and $\theta_0 + 2\pi k$ are equivalent for any integer $k$. So, for example, we expect the phases $0.99\pi$ and $-0.99\pi$ to be close together, the difference between them being $\vert\sfracpart{0.99\pi + 0.99\pi}_\pi\vert = 0.02\pi$, and not $\vert 0.99\pi + 0.99\pi\vert = 1.98\pi$.  %In the literature it is usual to write \emph{estimator} minus \emph{error}, i.e., $\hat{\theta} - \theta_0$ rather than $\theta_0 - \hat{\theta}$.  However, it will be notationally convenient in Sections~\ref{sec:proof-almost-sure} and~\ref{sec:proof-asympt-norm} to use $\langle\theta_0 - \hat{\theta}\rangle_\pi$.  
The theorems require minor modification when the number of pilots symbols $\abs{P} = 0$.  Theorem statements for this special case appear in \cite{McKilliam_leastsqPSKnoncoICASSP_2012}.

Part~\ref{thm:consistency:part1} of Theorem~\ref{thm:consistency} shows that the phase estimator error $\langle\theta_0 - \hat{\theta}\rangle_\pi$ converges almost surely to zero as $L\rightarrow \infty$, i.e., the estimator $\hat{\theta}$ is strongly consistent. In contrast, part~\ref{thm:consistency:part2} of the theorem shows that $\hat{\rho}$ is not a consistent estimator of the true amplitude $\rho_0$ and converges to $\rho_0 G(0)$.  The value $G(0)$ is a function of the noise distribution and $G(0)$ is close to $1$ when the variance of the noise is small, but grows without bound as the noise variance increases.  Thus, the asymptotic bias is small when the SNR is large, but significant when the SNR is small.  We will further investigate the bias in Section~\ref{sec:simulations}.  Theorem~\ref{thm:normality} shows that $\sqrt{L}(\hat{\rho} - \rho_0 G(0))$ (the amplitude estimator error with bias removed) and $\sqrt{L}\langle\theta_0 - \hat{\theta}\rangle_\pi$ are asymptotically normally distributed and statistically independent as $L\rightarrow\infty$.

@@

Recall the functions $G$, $h_1$ and $h_2$ defined in the statement of Theorem~\ref{thm:consistency}.  Observe that
\[
\expect G_L(\lambda) = \frac{\abs{P}}{L} h_1(\lambda) + \frac{\abs{D}}{L} h_2(\lambda)
\]
and since $\frac{\abs{P}}{L} \rightarrow p$ and $\frac{\abs{D}}{L} \rightarrow d$ as $L \rightarrow \infty$,
\[
\lim_{L \rightarrow\infty} \expect G_L(\lambda) = G(\lambda) = p h_1(\lambda)   +  d h_2(\lambda).
\]
%This is kind of disjointed without the ULL included.  Unfortunately there is no space.
As is customary, let $\Omega$ be the sample space on which the random variables $\{w_i\}$ are defined.  Let $A$ be the subset of the sample space $\Omega$ on which $G(\hat{\lambda}_L) \rightarrow G(0)$ as $L\rightarrow\infty$.  Lemma~\ref{lem:convtoexpGlamL} shows that $\prob\{A\} = 1$.  Let $A'$ be the subset of the sample space on which $\hat{\lambda}_L \rightarrow 0$ as $L\rightarrow \infty$.  Because $G(x)$ is uniquely maximised at $x=0$, it follows that $G(\hat{\lambda}_L) \rightarrow G(0)$ only if $\hat{\lambda}_L \rightarrow 0$ as $L \rightarrow\infty$. So $A \subseteq A'$ and therefore $\prob\{A'\} \geq \prob\{A\} = 1$.  Part 1 of Theorem~\ref{thm:consistency} follows.

@@

Figures~\ref{fig:plotphaseBPSK},~\ref{fig:plotphaseQPSK}~and~\ref{fig:plotphase8PSK} show the sample mean square error (MSE) of the phase estimator when $M=2,4,8$ with $L=4096$ and for varying proportions of pilots symbols $\abs{P} = 0, \tfrac{L}{32}, \tfrac{L}{8}, \tfrac{L}{2}, L$.  When $\abs{P} \neq 0$ the mean square error is computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}_\pi^2$.  Otherwise, when $\abs{P}=0$ the mean square error is computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}^2$ as in~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  The dots, squares, circles and crosses are the results of Monte-Carlo simulations with the least square estimator.  The solid lines are the estimator MSEs predicted by Theorem~\ref{thm:normality}.   %(and by Theorem~2 from~\cite{McKilliam_leastsqPSKnoncoICASSP_2012} when $\abs{P} = 0$).  
The prediction is made by dividing the asymptotic covariance matrix by $L$.  The theorem accurately predicts the behaviour of the phase estimator when $L$ is sufficiently large.  As the SNR decreases the variance of the phase estimator approaches that of the uniform distribution on $[-\pi, \pi)$ when $\abs{P} \neq 0$, and the uniform distribution on $[-\tfrac{\pi}{M}, \tfrac{\pi}{M})$ when $\abs{P}=0$~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  As the SNR increases the variance of the estimators converge to that of the estimator where all symbols are pilots, i.e. $\abs{P} = L$.

>>>>

In this paper we present an alternative approach to the study of estimators for carrier phase and amplitude.  Rather than compute bounds, we directly study the least squares estimator asymptotically as the number of transmitted symbols $L$ approaches $\infty$.  We show, under comparatively mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent (asymptotically unbiased) and asymptotically normally distributed.  However, the least squares amplitude estimator is asymptotically biased and converges to a positive real number that is a function of the true carrier amplitude, the noise distribution, and the size of the constellation $M$.  To our knowledge, this is the first time an analytic description of the bias or variance of a non-data-aided amplitude estimator has been obtained.  Our expressions for the asymptotic variance and bias are simple to compute.  %Our results are indirectly supported by the simulation study of Pauluzzi~\cite{Pauluzzi2000}.  We present our our set of simulations supporting our analytic results in Section~\ref{}.

@@

Turbo synchronisation can require a large number of EM iterations.  For applications with low power receivers, such as sensor networks, the complexity of turbo synchronisation might be too great and computationally simple channel estimators are required.  For example, in practice it is common to rely on channel estimates obtained from pilot symbols alone.  In this case the number of pilot symbols must be sufficiently large for channel estimates to be sufficiently accurate.  If more accuracy is required, then more pilot symbols must be used.  Our estimator provides a simple mechanism for increasing accuracy without increasing the number pilot symbols.  Instead, the pilot symbols and a subset of the data symbols can be jointly used for channel estimation.  %For computational complexity reasons it may not be desirable to use all of the existing data symbols for channel estimation.  This is easily possible with our estimator by choosing the indices in $D$ to correspond with the subset of the data symbols that are to be used.

@@

Define the sum of squares function
\begin{align}
LS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \nonumber \\
&= \sum_{i \in P \cup D} \big( \abs{y_i}^2 - a s_i y_i^* - a^* s_i^* y_i + aa^* \big), \label{eq:LS}
\end{align}
where $*$ denotes complex conjugate.  The minimiser of $LS$ with respect to $a$ as a function of $\{d_i, i \in D\}$ is
\begin{equation}\label{eq:hata}
\hat{a}(\{d_i, i \in D\}) = \frac{1}{L} \sum_{i \in P \cup D} y_i s_i^* = \frac{1}{L} Y,
\end{equation}
where $L = \abs{P \cup D}$ and to simplify notation we have put 
\[
Y = \sum_{i \in P \cup D} y_i s_i^* = \sum_{i \in P } y_i p_i^* + \sum_{i \in D } y_i d_i^*.
\]  
Note that $Y$ is a function of the unknown data symbols $\{ d_i, i \in D\}$ and we could write $Y(\{ d_i, i \in D\})$, but have chosen to suppress the argument $(\{ d_i, i \in D\})$ for notational brevity.  Substituting $\frac{1}{L}Y$ for $a$ into~\eqref{eq:LS} we obtain $LS$ minimised with respect to $a$,
\begin{equation}\label{eq:LSdatasymbols}
LS(\{d_i, i \in D\}) = A - \frac{1}{L}\abs{Y}^2,
\end{equation}
where $A = \sum_{i \in P \cup D}\abs{y_i}^2$ does not depend on the data symbols.  The least squares estimators of the data symbols are the minimisers of~\eqref{eq:LSdatasymbols}.  Observe that given candidate values for the data symbols, we can compute the corresponding $LS(\{d_i, i \in D\})$ in $O(L)$ arithmetic operations.  It turns out that there are at most $M\abs{D}$ candidate values for the least squares estimator of the data symbols~\cite{Sweldens2001,Mackenthun1994}.  %When the number of data symbols is not small, this set is substantially smaller than the entire set of possible transmitted symbols, which is of size $M^{\abs{D}}$.

To see this, let $a = \rho e^{j\theta}$ where $\rho > 0$ is real.  Now $LS$ from~\eqref{eq:LS} is
\begin{align}
LS(\rho, \theta, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - \rho e^{j\theta} s_i }^2  \nonumber \\
&= \sum_{i \in P} \abs{ y_i - \rho e^{j\theta} p_i }^2 + \sum_{i \in D} \abs{ y_i - \rho e^{j\theta} d_i }^2. \label{eq:LSallparams}
\end{align}
We have slightly abused notation here by reusing $LS$. This should not cause confusion as $LS(a, \{d_i, i \in D\})$, $LS(\rho, \theta, \{d_i, i \in D\})$, and $LS(\{d_i, i \in D\})$ are easily told apart by their arguments.  For given $\theta$ and $\rho$, the least squares estimator of the $i$th data symbol $d_i$ is given by minimising $\abs{ y_i - \rho e^{j\theta} d_i }^2$, that is,
\begin{equation}\label{eq:hatdfinxtheta}
\hat{d}_i(\theta) = e^{j\hat{u}_i(\theta)} \qquad \text{where} \qquad \hat{u}_i(\theta) = \round{\angle( e^{-j\theta}y_i)},
\end{equation}
where $\angle(\cdot)$ denotes the complex argument (or phase), and $\round{\cdot}$ rounds its argument to the nearest multiple of $\frac{2\pi}{M}$.  A word of caution, the notation $\round{\cdot}$ is often used to denote rounding to the nearest \emph{integer}.  This is not the case here.  If the function $\operatorname{round}(\cdot)$ takes its argument to the nearest integer then,
\[
\round{x} = \tfrac{2\pi}{M}\operatorname{round}\left(\tfrac{M}{2\pi}x\right).
\] 
It is common to call $\hat{d}_i(\theta)$ a `hard decision'  made for the $i$th data symbol under the hypothesis that the channel phase is $\theta$.  Note that $\hat{d}_i(\theta)$ does not depend on $\rho$.  As defined, $\hat{u}_i(\theta)$ is not strictly inside the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$, but this is not of consequence, as we intend its value to be considered equivalent modulo $2\pi$.  With this in mind,
\[
\hat{u}_i(\theta) = \round{\angle{y_i} - \theta }
\]
which is equivalent to the definition from~\eqref{eq:hatdfinxtheta} modulo $2\pi$.

@@

Let,
\begin{equation}\label{eq:LSfk}
LS(f_k) = A - \frac{1}{L}\abs{Y_k}^2,
\end{equation}
where, 
\[
Y_k = Y( f_k ) = \sum_{i \in P} y_i p_i^*  + \sum_{i \in D} y_i f_{k,i}^* = B + \sum_{i \in D}g_{k,i},
\]
where $B = \sum_{i \in P} y_i p_i^*$ is independent of the data symbols, and $f_{k,i}$ denotes the $i$th symbol in $f_k$ and, for convenience, we put $g_{k,i}  = y_i f_{k,i}^*$.  Letting $g_{k}$ be the sequence $\{g_{k,i}, i \in D\}$ we have, from~\eqref{eq:fkrec}, that $g_k$ satisfies the recursive equation
\[
g_{k+1} = g_{k} e_{\sigma(k)}^*,
\]
where $g_{k} e_{\sigma(k)}^*$ indicates the sequence $g_k$ with the $\sigma(k)$th element replaced by $g_{k,\sigma(k)}e^{j2\pi/M}$.  Now,
\[
Y_0 = B + \sum_{i \in D} g_{0,i}
\] 
can be computed in $O(L)$ operations, and
\begin{align*}
Y_1 &= B + \sum_{i \in D} g_{1,i} \\
&= B +  (e^{j2\pi/M} - 1)g_{0,\sigma(0)} + \sum_{i \in D} g_{0,i} \\
&= Y_0 + \eta g_{0,\sigma(0)},
\end{align*}
where $\eta = e^{j2\pi/M} - 1$.  In general,
\[
Y_{k+1} = Y_k + \eta g_{k,\sigma(k)}.
\]
So, each $Y_k$ can be computed from its predecessor $Y_{k-1}$ in $O(1)$ arithmetic operations.  Given $Y_k$, the value of $LS(f_k)$ can be computed in $O(1)$ operations using~\eqref{eq:LSfk}.  Let $\hat{k} = \arg\min LS(f_k)$.  The least squares estimator of $a_0$ is then computed according to~\eqref{eq:hata}, that is,
\[
\hat{a} = \frac{1}{L} Y_{\hat{k}}.
\]
Pseudocode is given in Algorithm~\ref{alg:loglinear}.  Line~\ref{alg_sortindices} contains the function $\operatorname{sortindices}$ that, given $z = \{z_i, i \in D\}$, returns the permutation $\sigma$ as described in~\eqref{eq:sigmasortind}.  The $\operatorname{sortindices}$ function requires sorting $\abs{D}$ elements.  This requires $O(L \log L)$ operations.  The $\operatorname{sortindices}$ function is the primary bottleneck in this algorithm when $L$ is large.  The loops on lines~\ref{alg_loop_setup} and~\ref{alg_loop_search} and the operations on lines~\ref{alg_Y} to lines~\ref{alg_Q} all require $O(L)$ or less operations.  %In the next sections we will show how to the sortinces function can be avoided.  This leads to an algorithm that requires only $O(L)$ operations.

@@

\begin{theorem}\label{thm:consistency} (Almost sure convergence)
Let $\{w_i\}$ be a sequence of independent and identically distributed, circularly symmetric complex random variables with $\expect \abs{w_1}^2$ finite, and let $\{y_i, i \in P \cup D\}$ be given by~\eqref{eq:sigmod}.   Let $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ be the least squares estimator of $a_0 = \rho_0e^{j\theta_0}$. %where both $\hat{\rho}$ and $\rho_0$ are positive real numbers.  
Put $L = \abs{P \cup D}$ and let $\abs{P}$ and $\abs{D}$ increase in such a way that
\[
\frac{\abs{P}}{L} \rightarrow p \qquad \text{and} \qquad \frac{\abs{D}}{L} \rightarrow d \qquad \text{as $L \rightarrow \infty$.}
\] 
Let $R_i \geq 0$ and $\Phi_i \in [0,2\pi)$ be real random variables satisfying
\begin{equation}\label{eq:RiandPhii}
R_ie^{j\Phi_i} =  \frac{a_0s_i + w_i}{a_0s_i}  = 1 + \frac{w_i}{a_0 s_i} ,
\end{equation}
and define the continuous function
\[
G(x) = p h_1(x) + d h_2(x) \qquad \text{where}
\]
\[
h_1(x) = \expect R_1 \cos(x + \Phi_1), \;\;\; h_2(x) =  \expect R_1 \cos\sfracpart{ x + \Phi_1}.
\]
If $p > 0$ and if $G(x)$ is uniquely maximised at $x = 0$ over the interval $[-\pi,\pi)$ then
\begin{enumerate}
\item $\sfracpart{\theta_0 - \hat{\theta}}_\pi \rightarrow 0$ almost surely as $L \rightarrow \infty$, \label{thm:consistency:part1}
\item $\hat{\rho} \rightarrow \rho_0 G(0)$ almost surely as $L \rightarrow \infty$. \label{thm:consistency:part2}
\end{enumerate}
\end{theorem}

@@

The proof of Theorem~\ref{thm:consistency} is in Section~\ref{sec:proof-almost-sure} and the proof of Theorem~\ref{thm:normality} is in Section~\ref{sec:proof-asympt-norm}. Before giving the proofs we discuss some of the consequences and the assumptions made by these theorems.  
First observe that the theorems place conditions on $\sfracpart{\theta_0 - \hat{\theta}}_\pi$ rather than directly on $\theta_0 - \hat{\theta}$.  This makes sense because the phases $\theta_0$ and $\theta_0 + 2\pi k$ are equivalent for any integer $k$. So, for example, we expect the phases $0.99\pi$ and $-0.99\pi$ to be close together, the difference between them being $\vert\sfracpart{0.99\pi + 0.99\pi}_\pi\vert = 0.02\pi$, and not $\vert 0.99\pi + 0.99\pi\vert = 1.98\pi$.  %In the literature it is usual to write \emph{estimator} minus \emph{error}, i.e., $\hat{\theta} - \theta_0$ rather than $\theta_0 - \hat{\theta}$.  However, it will be notationally convenient in Sections~\ref{sec:proof-almost-sure} and~\ref{sec:proof-asympt-norm} to use $\langle\theta_0 - \hat{\theta}\rangle_\pi$.  
The theorems require minor modification when the number of pilot symbols $\abs{P} = 0$.  Theorem statements for this special case appear in \cite{McKilliam_leastsqPSKnoncoICASSP_2012}.

Part~\ref{thm:consistency:part1} of Theorem~\ref{thm:consistency} shows that the phase estimator error $\langle\theta_0 - \hat{\theta}\rangle_\pi$ converges almost surely to zero as $L\rightarrow \infty$, i.e., the estimator $\hat{\theta}$ is strongly consistent. In contrast, part~\ref{thm:consistency:part2} of the theorem shows that $\hat{\rho}$ is not a consistent estimator of the true amplitude $\rho_0$ and converges to $\rho_0 G(0)$.  The value $G(0)$ is a function of the noise distribution and $G(0)$ is close to $1$ when the variance of the noise is small, but grows without bound as the noise variance increases.  Thus, the asymptotic bias is small when the SNR is large, but significant when the SNR is small.  We will further investigate the bias in Section~\ref{sec:simulations}.  Theorem~\ref{thm:normality} shows that $\sqrt{L}(\hat{\rho} - \rho_0 G(0))$ and $\sqrt{L}\langle\theta_0 - \hat{\theta}\rangle_\pi$ are asymptotically normally distributed and statistically independent as $L\rightarrow\infty$.

@@

Recall the functions $G$, $h_1$ and $h_2$ defined in the statement of Theorem~\ref{thm:consistency}.  Observe that
\[
\expect G_L(\lambda) = \frac{\abs{P}}{L} h_1(\lambda) + \frac{\abs{D}}{L} h_2(\lambda)
\]
and since $\frac{\abs{P}}{L} \rightarrow p$ and $\frac{\abs{D}}{L} \rightarrow d$ as $L \rightarrow \infty$,
\[
\lim_{L \rightarrow\infty} \expect G_L(\lambda) = G(\lambda) = p h_1(\lambda)   +  d h_2(\lambda).
\]
%This is kind of disjointed without the ULL included.  Unfortunately there is no space.
As is customary, let $\Omega$ be the sample space on which the random variables $\{w_i\}$ are defined.  Let $\Sigma$ be the subset of the sample space $\Omega$ on which $G(\hat{\lambda}_L) \rightarrow G(0)$ as $L\rightarrow\infty$.  Lemma~\ref{lem:convtoexpGlamL} shows that $\prob\{\Sigma\} = 1$.  Let $\Sigma'$ be the subset of the sample space on which $\hat{\lambda}_L \rightarrow 0$ as $L\rightarrow \infty$.  Because $G(x)$ is uniquely maximised at $x=0$, it follows that $G(\hat{\lambda}_L) \rightarrow G(0)$ only if $\hat{\lambda}_L \rightarrow 0$ as $L \rightarrow\infty$. So $\Sigma \subseteq \Sigma'$ and therefore $\prob\{\Sigma'\} \geq \prob\{\Sigma\} = 1$.  Part 1 of Theorem~\ref{thm:consistency} follows.

@@

Figures~\ref{fig:plotphaseBPSK},~\ref{fig:plotphaseQPSK}~and~\ref{fig:plotphase8PSK} show the sample mean square error (MSE) of the phase estimator when $M=2,4,8$ with $L=4096$ and for varying proportions of pilot symbols $\abs{P} = 0, \tfrac{L}{32}, \tfrac{L}{8}, \tfrac{L}{2}, L$.  When $\abs{P} \neq 0$ the mean square error is computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}_\pi^2$.  Otherwise, when $\abs{P}=0$ the mean square error is computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}^2$ as in~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  The dots, squares, circles and crosses are the results of Monte-Carlo simulations with the least square estimator.  The solid lines are the estimator MSEs predicted by Theorem~\ref{thm:normality}.   %(and by Theorem~2 from~\cite{McKilliam_leastsqPSKnoncoICASSP_2012} when $\abs{P} = 0$).  
The prediction is made by dividing the asymptotic covariance matrix by $L$.  The theorem accurately predicts the behaviour of the phase estimator when $L$ is sufficiently large.  As the SNR decreases the variance of the phase estimator approaches that of the uniform distribution on $[-\pi, \pi)$ when $\abs{P} \neq 0$, and the uniform distribution on $[-\tfrac{\pi}{M}, \tfrac{\pi}{M})$ when $\abs{P}=0$~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  As the SNR increases the variance of the estimators converge to that of the estimator where all symbols are pilots, i.e. $\abs{P} = L$.
